### 2016-06-16 motivation.tex
Finally motivated!

Firstly, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, hence admits \emph{rely-guarantee} style
reasoning~\cite{rgjones}. Rely-guarantee is compositional proof
technique that allows us to reason about one thread at a time by
abstracting away the interference due to remaining threads
(collectively called \emph{the environment}) into a \emph{rely}
relation. In an ordinary concurrent program, every environment step is
a valid interference in the current thread. However, in presence of
transactions and variable isolation, determining what constitutes an
interference and what does not is a non-trivial problem. For example,
inside a serializable transaction no interference is a valid
interference, whereas inside a transaction executing under \iso{Read
Uncommitted}, the weakest isolation level, all interferences are
valid. Between these two extremes are various levels of isolation
that admit some interferences while prohibiting others. 
% For example, \iso{Read Committed} isolation admits interference of a
% committed transaction, but not that of an uncommitted transaction.
% \iso{Snapshot Isolation} admits interference from committed
% non-conflicting transactions, and so on. 
The key aspect of our proof system is that it facilitates
rely-guarantee reasoning while allowing the programmer to ignore any
interference that is invalid as per the chosen isolation level or the
store consistency level. For instance, in the current example, `Wd2'
executes concurrently with `Wd1', hence latter's rely relation is
required to model all possible interferences from `Wd1'. However,
since RC prevents the effects of an uncommitted transaction from being
visible, our proof system allows the programmer to effectively ignore
the interference from `Wd2' unless the interference includes its
commit. RC execution 1 shown above is an example of a case where
interference from `Wd2' includes its commit. Such interference is not
prohibited by RC, and the SC store makes \emph{all} effects of `Wd2'
immediately visible. Our proof system insists that the
programmer consider all effects of `Wd2' and conclude the value of
\C{C} as \C{k-a2}, thus leading (rightfully) to the failure of proof.
On the other hand, \iso{Snapshot Isolation} proscribes any
interference from a concurrent transaction performing conflicting
writes. The RC executions shown above are therefore invalid under SI.
Accordingly, programmer is allowed to ignore any interference from
`Wd2', thus facilitating the proof of postcondition.

As evident from the above example, a proof system for weak isolation
has to be flexible enough to admit different kinds of reasoning for
different isolation and consistency levels. One way this can be
achieved is by having a separate proof rule for each combination of
transaction isolation level and store consistency level, such that the
rule reflects the operational semantics of the isolation level on the
store. Given the large and ever-increasing diversity among isolation
and consistency semantics, this approach is unlikely to scale. Our
proof system avoids this problem by being parametric over the
semantics of isolation and consistency levels. This approach is based
on a couple of observations. First, as noted by some
authors~\cite{pldi15,gotsmanconcur15}, semantics of various isolation
and consistency levels can be captured as well-formedness conditions
over execution traces. If an execution trace satisfies specified
well-formedness conditions, then the program is considered to have
experienced the required levels of consistency and isolation in that
execution. Second, well-formedness conditions are often
\emph{prefix-complete}, meaning that every prefix of a well-formed
trace is also well-formed, or conversely, an ill-formed trace cannot
be extended to a well-formed one. Prefix-completeness allows us to
reason about programs operationally, in terms of an imaginary machine
whose primary artifact is an execution trace. The machine starts from
a well-formed trace, executes a read/write/commit operation, and
extends the trace while preserving well-formedness. If well-formedness
cannot be preserved by taking a step, the machine gets ``stuck''. In
executions that run to completion, the machine models a store whose
consistency and isolation guarantees are captured by the trace
well-formedness conditions. Thus, to show that program's invariants
are preserved under weak isolation, it suffices to show that they are
preserved in all executions of the imaginary machine that do not get
stuck. This is indeed how our proof system facilitates program
verification. The ``invalid interference'' described informally before
corresponds to a case of machine getting stuck due to the violation of
trace well-formedness conditions. The proof system inherits the
parametricity of the machine over trace well-formedness conditions,
leading to a generic set of parameterized proof rules that can be
instantiated for various store consistency and transaction isolation
levels.

We now demonstrate how rely-guarantee reasoning, parametricity ,and
trace well-formedness conditions facilitate the proof of the example
in Fig.~\ref{fig:motiv-eg-1}. The trace-wellformedness condition
corresponding to \iso{Snapshot Isolation} is shown below\footnote{This
is a simplified version of the actual specification of SI
(\S~\ref{sec:opsem}), which is a little more involved.}:
\begin{smathpar}
\begin{array}{lcl}
\underE{\C{SI}(T_j)} & \Rightarrow & \forall T_i. T_i \neq T_j \conj \\
  & & (\exists \C{X}.~\underE{T_i \wrstoar \C{X}} \conj 
                  \underE{T_j \wrstoar \C{X}}) \Rightarrow \\
  &  & \underE{T_i \hboar T_j} \disj \underE{T_j \hboar \C{COMMIT}(T_i)} \\
\end{array}
\end{smathpar}
A transaction $T_j$ that writes to a variable $\C{X}$ has experienced
snapshot isolation in an execution $\E$ only if every other
transaction $T_i$ that also wrote to \C{X} in $\E$ either happened
before (i.e., committed before) $T_j$, or executed concurrently but
commited after $T_j$. Here, $\E$ denotes an execution trace, and
$\underE{\phi}$ asserts that proposition $\phi$ about $\E$ is valid.
Since we require both `Wd1' and `Wd2' to be executed under SI, the
trace invariant ($\I$) for the example is $\C{SI(Wd1)} \wedge
\C{SI(Wd2)}$. That is, if the machine extends an execution $\E$ that
satisfies $\I$ (i.e.,$\underE{\I}$) to $\E'$, then $\E'$ also satisfies
$\I$ (i.e., $\E' \Vdash \I$). Since we reason in terms of traces, pre
and post conditions need to be written with respect to a trace.

\begin{figure}
\centering
\begin{txnimpcode}
 $\begin{decoration}
 P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
        \hboar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
 \end{decoration}$
  txn$\langle$'Wd1'$\rangle${
   $\begin{decoration}
    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \hboar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \}
    \end{decoration}$ 
    v1 = C
   $\begin{decoration}
    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \hboar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
    \end{decoration}$ 
    if (v1 $\ge$ a1) {
      v2 = C;
     $\begin{decoration}
      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \hboar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
          \hspace*{1.9in}\C{v2 = k-a2}\}
      \end{decoration}$ 
      v3 = v2 - a1;
     $\begin{decoration}
      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \hboar \C{Wd1}} 
         \Rightarrow \C{C = k-a2} \wedge \\
         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
      \end{decoration}$ 
      C := v3
     $\begin{decoration}
      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
                \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} 
                \Rightarrow \C{C = k-a1-a2}\}
      \end{decoration}$ 
    }
  }
 $\begin{decoration}
  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
            \conj \committed(\C{Wd1}) \conj\\
      \hspace*{0.12in}{\committed(\C{Wd2})} 
          \Rightarrow \C{C = k-a1-a2} \}
  \end{decoration}$ 
\end{txnimpcode}

\caption{`Wd1' transaction decorated with assertions}
\label{fig:wd1-decorated}
\end{figure}

The fully decorated implementation of transaction `Wd1' is shown in
Fig.~\ref{fig:wd1-decorated}. Decorated `Wd2' is similar. Assignment
statements are broken down and temporary local variables (\C{v1},
\C{v2} and \C{v3}) are introduced so as to separate shared variable
reads and writes.  Since we reason in terms of executions, all
assertions implicitly refer to the current execution (just as hoare
triples implicitly refer to the current state). Also implicit is the
invariant \C{k $\ge$ a1+a2}. Proposition $\committed(T)$ indicates
that the transaction $T$ is already committed. Precondition ($P_1$) of
`Wd1' accounts for the possibility of `Wd2' committing before `Wd1',
writing \C{k-a2} to \C{C}.  Precondition of `Wd2' is similar.  Since
neither `Wd1' nor `Wd2' are committed at the beginning, the
precondition in Fig.~\ref{fig:motiv-eg-1} is extended to include
$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
$P_1$ follows. Once the execution is inside `Wd1', commit of `Wd2'
($\committed(\C{Wd2})$) may mean that either `Wd2' happened before
`Wd1', or that it committed concurrently with `Wd1'. Since SI
proscribes the latter possibility, we only consider the case when
$\C{Wd2} \hboar \C{Wd1}$. A proof for $\C{Wd2} \hboar \C{Wd1}$ is
obtained subsequently, allowing us to get rid of the special case.

The first obligation is to show that the assertions that decorate
`Wd1' are indeed valid. This entails proving that, for each statement
of `Wd1', if evaluating the statement and extending an execution that
satisfies its precondition results in a well-formed execution, then
the extended execution must satisfy the postcondition. The
well-formedness assumption lets us ignore the executions that violate
the trace invariant lead the machine to a stuck state. This is useful
in case of the assignment statement that writes \C{v3} to \C{C}. Since
`Wd2' also writes to \C{C}, the trace invariant ($\I$) requires
$\C{Wd2} \hboar \C{Wd1}$ if $\committed(\C{Wd2})$ (the alternative
$\C{Wd1} \hboar \C{COMMIT}(\C{Wd2})$ allowed by \C{SI(Wd1)} is
impossible in this case because \C{Wd2} has already committed, while
\C{Wd1} is still in progress). This lets us treat the special case
$\committed(\C{Wd2})\conj \C{Wd2} \hboar \C{Wd1}$ as being equivalent
to the general case $\committed(\C{Wd2})$, which is required to prove
the postcondition of `Wd1'. Observe that \C{RC(Wd1)} does not justify
this generalization, hence replacing \C{SI(Wd1)} by \C{RC(Wd1)} is not
sufficient to prove the postcondition of `Wd1'.

The second obligation is to prove that assertions are valid despite
the interference from the the concurrent thread executing `Wd2'. The
interference is given by the following rely relation ($R_1$):
\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
        \underE{\I} \conj \E'\Vdash \I \conj\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
\end{array}
\end{smathpar}
The rely relation says the following about the concurrent thread: (1).
It may interfere only if `Wd2' is not already committed, (2). Any
interference takes well-formed executions to well-formed executions,
(3). If the interference commits `Wd2', then `Wd2' should have already
written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
or $\C{k-a1}$ depending on whether or not `Wd1' has already committed.
Interference from `Wd2' is allowed between any two operations of
`Wd1', and assertions are required to be valid despite the
interference. For instance, if the execution ($\E$) immediately after
evaluating $\C{v1 = C}$ satisfies $\phi_2$ (i.e., $\underE{\phi_2}$),
and an interference from `Wd2' extends $\E$ to $\E'$ (i.e.,
$R_1(\E,\E')$), then the execution $\E'$ just before evaluating the
condition must satisfy $\phi_2$ (i.e., $\E' \Vdash \phi_2$). Indeed,
all $\phi_i$'s do satisfy this \emph{stability} condition, thus
lending the validity to the proof.

>>>>

\begin{figure}
\centering
  \includegraphics[scale=0.4]{Figures/motiv-eg-1-hb}

\label{fig:motiv-eg-1-hb}
\caption{$\hbZ$ arrows allowed by SI (solid) and RC (solid and dashed)
isolation levels for the example in Fig.~\ref{fig:motiv-eg-1}. Each
arrow represents $\hbZ$ relationship between all operations of `Wd2'
and operations in `Wd1' that follow the arrow head. Letters R, W and C
stand for read, write and commit, respectively.}
\end{figure}

Firstly, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, hence admits \emph{rely-guarantee} style
reasoning~\cite{rgjones}. Rely-guarantee is compositional proof
technique that allows us to reason about one thread at a time by
abstracting away the interference due to remaining threads
(collectively called \emph{the environment}) into a \emph{rely}
relation. In an ordinary concurrent program, every environment step is
a valid interference in the current thread. However, in presence of
transactions and variable isolation, determining what constitutes an
interference and what does not is a non-trivial problem. For example,
inside a serializable transaction no interference is a valid
interference, whereas inside a transaction executing under \iso{Read
Uncommitted}, the weakest isolation level, all interferences are
valid. Between these two extremes are various levels of isolation
that admit some interferences while prohibiting others. 
% For example, \iso{Read Committed} isolation admits interference of a
% committed transaction, but not that of an uncommitted transaction.
% \iso{Snapshot Isolation} admits interference from committed
% non-conflicting transactions, and so on. 
For the rely guarantee approach to be useful it has to constrain the
interference in accordance with the chosen level of isolation. The
first key observation we make is that we can constrain interference by
constraining the nature of \emph{happens-before} ($\hbZ$) relation
between transactions or their respective constituents. We make use of
the $\hbZ$ relation to axiomatize the interference characteristic of
isolation levels as their specifications ($\psi$). For example,
\iso{Read Committed} specification allows an operation $\eta_1$ of
$T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
operation in $T_1$ (including its commit) happens before $\eta_2$ (we
write $T_1 \hboar \eta_2$ to denote this). RC spec does not require
$T_1$ to happen before all operations of $T_2$, thus allowing it to
interfere in $T_2$, as possible under RC isolation. \iso{Snapshot
Isolation} spec requires two transactions that write to the same
variable to be related by happens-before.  This effectively proscribes
interference due to $T_1$ in $T_2$ (or, $T_2$ in $T_1$).
Formally\footnote{Note that specs are simplified for illustration. The
actual specs (\S~\ref{sec:opsem}) are more nuanced.  }:
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall \eta_1,\eta_2,T_1,T_2.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ arrows
from `Wd2' to `Wd1' allowed by $\psi_{RC}$and $\psi_{SI}$. All
arrows re legal under $\psi_{RC}$ because in no case does an operation
$\eta_1$ from `Wd1' happen before $\eta_2$ of `Wd2' without the commit
of `Wd1' also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, becase dashed arrows establish $\hbZ$ between `Wd2'
and only a subset of operations in `Wd1'. Ligher (blue) arrows with
hollow heads denote $\hbZ$ relationships that do not effect the value
of \C{C} in a way that causes the program to violate its
postcondition.  Darker (black) arrows with solid heads denote $\hbZ$
relationships that lead to the violation of postcondition. The task of
the reasoning framework is to determine if all $\hbZ$ relationships
allowed by an isolation level lead to the satisfaction of the
postcondition.

% There is a $\hb$ edge (first dashed) that is allowed by RC but
% prohibited by SI, which nonetheless does not lead to invariant
% violation.

Specification of an isolation level encodes its interference
characteristic as constraints over $\hbZ$ relation. However, for this
to be useful in reasoning about programs, rely guarantee framework
should be able to use the specification to determine if an
interference is valid or invalid, allowing the programmer to only
focus on the former. Our second key observation is that this is
possible if the reasoning framework adequately tracks $\hbZ$ at each
program point, while preserving $\hbZ$ constraints as invariants
between program points. An interference that leads to the violation of
$\hbZ$ constraints (i.e., an invalid interference) is thus
automatically prohibited. For instance, consider the program point
after the write to \C{C} in `Wd1'. The expected invariant ($\phi$) at
that program point is shown below ($\committed$ stands for
``committed''):
\begin{smathpar}
\begin{array}{l}
  \neg\committed(\C{Wd1}) \conj (\neg\committed(\C{Wd2}) \Rightarrow
  \C{C = k-a1}) 
                \\
      \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
                \Rightarrow \C{C = k-a1-a2})\}
\end{array}
\end{smathpar}
$\phi$ asserts that `Wd1' is not yet committed, and that it wrote to
\C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}
depending on whether or not `Wd2' is committed. If $\phi$ stays
invariant until `Wd1' commits, then the postcondition (\C{C =
k-a1-a2}) can be established easily. However, an interference from
`Wd2' at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a1}$ to \C{C} and committing `Wd2', thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
proved by showing that an interference from `Wd2' starting from a
execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds (former does not hold because at least one write
from `Wd1' has happened before at least one operation of `Wd2', and
the latter doesn't hold because `Wd1' hasn't yet committed and `Wd2'
has already begun). Since `Wd2' also writes to \C{C}, this violates
the $\psi_{SI}$ constraint which we assume to be an invariant. A proof
for postcondition now follows from the contradiction. It is
informative to note that if the invariant is $\psi_{RC}$ instead of
$\psi_{SI}$, we cannot derive a contradiction and we cannot rule out
the interference, which causes (rightfully) the proof to fail.

We have thus far assumed an SC store that, in the absence of
transactions with special isolation requirements, makes effects of an
operation to be immediately visible to subsequent operations. Thus,
the natural behavior of SC store to totally order all operations w.r.t
$\hbZ$ relation could be in conflict with the constraints imposed by
weak isolation. To meet the requirements of weaker isolation levels,
say RC, the store has to adapt itself to \emph{hide} the effects of
concurrent transactions until they are committed. Once committed
though, the effects need to be immediately and perennially visible to
subsequent operations. This semantics of an SC store can be built into
the reasoning framework leading to a proof system tailor-made for such
stores. However, for the reasoning framework to be useful across the
stores, it has to be parametric over store consistency semantics, and
should be able to reconcile conflicts betweenconsistency and isolation
constraints. Later sections demonstrate how our reasoning framework
makes this possible.
